04_analysis

### 2023-10-20 05_analysis.Rmd
Update 05_analysis.Rmd
new rolling averages plots

---
title: "04_analysis"
author: "Marina Bozinovic"
date: "2023-05-11"
output: html_document
---
# Exploratory Analysis and Modeling for all 13 buoys

@@

# Create a new column that specifies the drift number before the bind.
for (col in dlist){
  data <- get(col) %>% mutate(Deployment = substr(col,17,18))
  assign(paste0("GPSwhaleAISenvSS", substr(col,17,18)), data)
}

# Bind and save allDrifts dataframe
allDrifts <- rbind(GPSwhaleAISenvSS07, GPSwhaleAISenvSS08, GPSwhaleAISenvSS10, GPSwhaleAISenvSS12,
                   GPSwhaleAISenvSS13, GPSwhaleAISenvSS14, GPSwhaleAISenvSS16, GPSwhaleAISenvSS18,
                   GPSwhaleAISenvSS19, GPSwhaleAISenvSS20, GPSwhaleAISenvSS21, GPSwhaleAISenvSS22,
                   GPSwhaleAISenvSS23)
save(allDrifts, file = "data/allDrifts.rda")
```

@@

ggplot(hourly, aes(x = time, y = mean_TOL2000)) + geom_line() + 
  geom_smooth() + 
  geom_point() +
  #geom_point(aes(color = species)) +
  #geom_vline(xintercept = hourly$UTC[hourly$shipin10km == 1], color = "yellow", alpha = 0.5) +
  scale_color_discrete(na.translate=FALSE) #+
 # scale_x_datetime(date_breaks = "1 day",limits = as.POSIXct(c("2018-09-24 00:00:00", "2018-10-03 00:00:00")))
```
# Sound levels across the whole range
```{r fig.width = 10}
ggplot(allDrifts, aes(UTC,TOL_2000)) + geom_point() 
```

# Using rollApply for ZC
```{r}
ZC_filt <- GPSwhaleAISenvSS23 %>% 
  filter(species %in% c("ZC", NA)) %>%
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA))

@@

#Plot facet grid
ggplot(ZC_pivot, aes(x = UTC, y = value, color = avg)) + geom_smooth() + 
  facet_grid(vars(avg), scales = "free") +
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 23")
```
# Using rollApply for PM
```{r}
PM_filt <- GPSwhaleAISenvSS23 %>% 
  filter(species %in% c("PM", NA)) %>%
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA))

@@

#Plot facet grid
ggplot(PM_pivot, aes(x = UTC, y = value, color = avg)) + geom_smooth() + 
  facet_grid(vars(avg), scales = "free") +
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Averages of PM Detections alongside Soundscape Metrics - Buoy 23")
```

Compare overall TOL2000sound levels when whales are absent vs. present
```{r}
ggplot(allDrifts, aes(factor(Wpresence), TOL_125)) + geom_boxplot()
```

@@

# All buoys
detsOnly3 <- allDrifts %>% filter(Wpresence == 1)
shipOnly3 <- allDrifts %>% filter(shipin10km == 1)
ggplot(detsOnly3) +
  geom_vline(xintercept = shipOnly3$UTC, color = "green") +  # add line for ship presence
  geom_point(aes(x = UTC, y = TOL_125, color = species), size = 2) +
  scale_x_datetime(date_breaks = "2 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#Sound levels for all species
```{r}
# Sound levels for all species
ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_125)) +
  labs(title = "TOL 125 Sound levels by Species - All Buoys")
ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_2000)) +
    labs(title = "TOL 2000 Sound levels by Species - All Buoys")
```

@@

#Hours After Ship Event by Species
```{r}
no_na <- allDrifts %>% filter(Wpresence == 1) %>%
  mutate(HrsafterShip = timeAfterShip /3600)
ggplot(no_na) + geom_histogram(aes(x = HrsafterShip), bins = 30) + facet_grid(species ~.) +
  labs(title = "Hours Elapsed from Ship Event by Species - All buoys")
```

>>>>

---
title: "04_analysis"
author: "Marina Bozinovic"
date: "2023-05-11"
output: html_document
---
# Exploratory Analysis for all 13 buoys

@@

# Create a new column that specifies the drift number before the bind.
for (col in dlist){
  data <- get(col) %>% mutate(Deployment = substr(col,17,18))
  data <- data[data$species %in% c(NA, "ZC", "PM", "BB", "BW43"),]    # Remove BW37V and BWC from analysis as there are too few detections.
  assign(paste0("GPSwhaleAISenvSS", substr(col,17,18)), data)
}

# Bind and save allDrifts dataframe
allDrifts <- rbind(GPSwhaleAISenvSS07, GPSwhaleAISenvSS08, GPSwhaleAISenvSS10, GPSwhaleAISenvSS12,
                   GPSwhaleAISenvSS13, GPSwhaleAISenvSS14, GPSwhaleAISenvSS16, GPSwhaleAISenvSS18,
                   GPSwhaleAISenvSS19, GPSwhaleAISenvSS20, GPSwhaleAISenvSS21, GPSwhaleAISenvSS22,
                   GPSwhaleAISenvSS23)

save(allDrifts, file = "data/allDrifts.rda")
load(file = "data/allDrifts.rda")
```

# Which buoy was the loudest?
```{r loudest-buoy}
allDrifts %>%
  ggplot(mapping = aes(x = Deployment, y = TOL_2000)) + geom_boxplot() + theme_bw()

allDrifts %>%
  group_by(Deployment) %>%
  summarize(range(TOL_2000))
```

# Distribution of noise by drift
```{r}
ggplot(allDrifts, aes(x = Deployment, y = TOL_2000)) + geom_boxplot() + 
  labs(title = "Distribution of TOL2000 by Drift") +
  theme_classic()

# Where is the loudest sound level?
allDrifts[allDrifts$TOL_2000 == max(allDrifts$TOL_2000),]

# Where is the quietest sound level?
allDrifts[allDrifts$TOL_2000 == min(allDrifts$TOL_2000),]
```

@@

ggplot(hourly, aes(x = time, y = mean_TOL2000)) + geom_line() + 
  geom_smooth() + 
  geom_point() +
  #geom_point(aes(color = species)) +
  #geom_vline(xintercept = hourly$UTC[hourly$shipin10km == 1], color = "yellow", alpha = 0.5) +
  scale_color_discrete(na.translate=FALSE) #+
 # scale_x_datetime(date_breaks = "1 day",limits = as.POSIXct(c("2018-09-24 00:00:00", "2018-10-03 00:00:00")))
```
# Sound levels across the whole range
```{r fig.width = 10}
ggplot(allDrifts, aes(UTC,TOL_2000, color = Deployment)) + geom_smooth() 
```

# Using rollApply for ZC
```{r}
ZC_filt <- GPSwhaleAISenvSS21 %>% 
  filter(species %in% c("ZC", NA)) %>%
  # Hourly rolling averages
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
  # Drop first and last NAs that are created as part of averages
  drop_na(., rollAvg_Whl)

@@

#Plot facet grid
ggplot(ZC_pivot, aes(x = UTC, y = value, color = avg)) + geom_smooth() +
  facet_grid(vars(avg), scales = "free") +
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Hourly Averages of ZC Detections and Soundscape Metrics -  
       Buoy 21")
```
# rollapply for vessels
```{r}
ZC_filt_ship <- GPSwhaleAISenvSS08 %>% 
  filter(species %in% c("ZC", NA)) %>%
  # Hourly rolling averages
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
  mutate(rollAvg_ship = zoo::rollapply(shipin10km, 5, mean, fill = NA)) %>%
  # Drop first and last NAs that are created as part of averages
  drop_na(., rollAvg_Whl)

ZC_pivot_ship <- pivot_longer(ZC_filt_ship, cols = rollAvg_Whl:rollAvg_ship, names_to = "avg", values_to = "value")

#Plot facet grid
ggplot(ZC_pivot_ship, aes(x = UTC, y = value, color = avg)) + geom_smooth() +
  facet_grid(vars(avg), scales = "free") +
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Hourly Averages of ZC Detections + SS Metrics + Ships -  
       Buoy 08")
```

Vessel presence vs ZC presence
```{r}
ZC_filt_ship <- GPSwhaleAISenvSS23 %>% 
  filter(species %in% c("ZC", NA)) %>%
  # Hourly rolling averages
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA)) %>%
  mutate(rollAvg_ship = zoo::rollapply(shipin10km, 5, mean, fill = NA)) %>%
  # Drop first and last NAs that are created as part of averages
  drop_na(., rollAvg_Whl)

# One plot
ggplot(ZC_filt_ship, aes(x = UTC)) +
  geom_smooth(aes(y = rollAvg_ship, color = "ship presence")) +
  geom_smooth(aes(y = rollAvg_Whl, color = "whale")) +  
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Hourly Averages of ZC Detections and Ships -  
       Buoy 23")

# rollApply on same plot for ZC
```{r}
# Plot lines on same plot

min_first <- min(ZC_filt$rollAvg_Whl)
max_first <- max(ZC_filt$rollAvg_Whl)
min_second <- min(ZC_filt$rollAvg_2000)
max_second <- max(ZC_filt$rollAvg_2000)

# scale and shift variables calculated based on desired mins and maxes
scale = (max_second - min_second)/(max_first - min_first)
shift = min_first - min_second

# Function to scale secondary axis
scale_function <- function(x, scale, shift){
  return ((x)*scale - shift)
}

# Function to scale secondary variable values
inv_scale_function <- function(x, scale, shift){
  return ((x + shift)/scale)
}

# Plot two lines on same plot
ggplot(ZC_filt, aes(x = UTC)) +
  geom_smooth(aes(y = rollAvg_Whl, color = "ZC presence")) +
  geom_smooth(aes(y = inv_scale_function(rollAvg_2000, scale, shift), color = "TOL 2000")) +
  scale_x_datetime(date_breaks = "3 days") +
  scale_y_continuous(limits = c(min_first, max_first), 
                     sec.axis = sec_axis(~scale_function(., scale, shift), 
                                         name = "TOL 2000 - hourly rolling average")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 13")

########################################
## Manual scale
scale = 2000
ggplot(ZC_filt, aes(x = UTC)) +
  geom_smooth(aes(y = rollAvg_Whl, color = "ZC presence")) +
  geom_smooth(aes(y = rollAvg_2000/scale, color = "TOL 2000")) +
  scale_x_datetime(date_breaks = "3 days") +
  scale_y_continuous(#limits = c(0.0, 0.05),
                     sec.axis = sec_axis(~.* scale, name = "TOL 2000 - hourly rolling average")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Averages of ZC Detections alongside Soundscape Metrics - Buoy 13")

```
```{r}
ggplot(ZC_filt, aes(x = UTC)) +
  geom_smooth(aes(y = rollAvg_Whl+0.25, color = "ZC presence")) #+
  geom_smooth(aes(y = rollAvg_2000, color = "TOL 2000"))
```

# Using rollApply for PM
```{r}
PM_filt <- GPSwhaleAISenvSS14 %>% 
  filter(species %in% c("PM", NA)) %>%
  mutate(rollAvg_Whl = zoo::rollapply(Wpresence, 5, mean, fill = NA)) %>%
  mutate(rollAvg_2000 = zoo::rollapply(TOL_2000, 5, mean, fill = NA)) %>%
  mutate(rollAvg_125 = zoo::rollapply(TOL_125, 5, mean, fill = NA))

@@

#Plot facet grid
ggplot(PM_pivot, aes(x = UTC, y = value, color = avg)) + geom_smooth() + 
  facet_grid(vars(avg), scales = "free") +
  scale_x_datetime(date_breaks = "3 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Rolling Averages of PM Detections alongside Soundscape Metrics - Buoy 14")
```

Compare overall TOL2000sound levels when whales are absent vs. present
```{r}
ggplot(allDrifts, aes(factor(Wpresence), TOL_2000)) + geom_boxplot()
```

@@

# All buoys
detsOnly3 <- allDrifts %>% filter(Wpresence == 1)
shipOnly3 <- allDrifts %>% filter(shipin10km == 1)
ggplot(detsOnly3) +
  geom_vline(xintercept = shipOnly3$UTC, color = "green") +  # add line for ship presence
  geom_point(aes(x = UTC, y = TOL_2000, color = species), size = 2) +
  scale_x_datetime(date_breaks = "2 days") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#Sound levels for all species
```{r}
# Sound levels for all species
ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_125)) +
  labs(title = "TOL 125 Sound levels by Species - All Buoys") +
  theme_classic()
ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = TOL_2000)) +
    labs(title = "TOL 2000 Sound levels by Species - All Buoys") +
  theme_classic()

# TOL_2000
detsOnly3 %>%
  group_by(species) %>%
  summarize(median(TOL_2000))

# Depth
ggplot(detsOnly3) + geom_boxplot(aes(x = species, y = temp400)) +
  labs(title = "Depth by Species - All Buoys") +
  theme_classic()
```

@@

#Hours After Ship Event by Species
```{r}
allDrifts %>%
  group_by(Deployment) %>%
  summarize(n = sum(shipin10km))

no_na <- allDrifts %>% filter(Wpresence == 1) %>%
  mutate(HrsafterShip = timeAfterShip /3600) %>%
  mutate(DaysAfterShip = HrsafterShip / 24)

# Hours after ship event
ggplot(no_na) + geom_histogram(aes(x = HrsafterShip), bins = 60) + facet_grid(species ~.) +
  labs(title = "Hours Elapsed from Ship Event by Species - All buoys") +
  scale_x_continuous(breaks = seq(0,500, by = 50)) +
  theme_bw()

# Days after ship event
ggplot(no_na) + geom_histogram(aes(x = DaysAfterShip), bins = 60) + facet_grid(species ~.) +
  scale_x_continuous(breaks = seq(0,20, by = 1)) +
  labs(title = "Days Elapsed from Ship Event by Species - All buoys") +
  theme_bw()

no_na %>%
  drop_na(DaysAfterShip) %>%
  group_by(species) %>%
  summarize(median(DaysAfterShip))
```
